---
title: "Evaluation of Basketball Statistics"
author: "Blake Nelson, Nick Olech"
date: "2024-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description of Data Set

For our project we are combining two different data sets.  The first data frame has information for each player such as personal information and draft information like what position, draft year, and draft team.  The second dataframe gives year by year information for each player in the first data frame. This information includes salary, season, and team.

### Loading packages and data frames
```{r, message=FALSE, error=FALSE, warning=FALSE}
# Load required packages
library(ggplot2)
library(tidyverse)
library(dplyr)
library(scales)
```

```{r}
# Draft information Data Frame
df1 <-
  read.csv(
    "https://query.data.world/s/7fyub35dc6n4crgjosveptbc7uxeo3?dws=00000",
    header = TRUE,
    stringsAsFactors = FALSE
  )

# Salary information Data Frame
df2 <-
  read.csv(
    "https://query.data.world/s/6xhwdm7emdeglx4botceotj6udfevo?dws=00000",
    header = TRUE,
    stringsAsFactors = FALSE
  )
```

# First Data Cleaning Steps

```{r, warning=FALSE}
# Merge df1 (draft info) and df2 (salary statistics)
nba_stats <-
  merge(df2,
        df1,
        by.x = "player_id",
        by.y = "X_id",
        all.x = TRUE)

# Replaces empty values with NA
nba_stats <- nba_stats %>%
  mutate(across(where(is.character), ~ na_if(.x, "")))

# Convert data types
nba_stats <- nba_stats %>%
  mutate(
    height = as.numeric(sub("([0-9]+)-([0-9]+)", "\\1", height)) * 12 +
      as.numeric(sub("([0-9]+)-([0-9]+)", "\\2", height)),
    career_FG. = as.numeric(career_FG.),
    career_FT. = as.numeric(career_FT.),
    career_PER = as.numeric(career_PER),
    career_PTS = as.numeric(career_PTS),
    career_TRB = as.numeric(career_TRB),
    career_WS = as.numeric(career_WS),
    draft_pick = as.numeric(gsub("[^0-9]", "", draft_pick)),
    # Extract numeric values from draft pick
    weight = as.numeric(gsub("[^0-9.]", "", weight))  # Extract numeric values from weight
  )

# Drop missing values
nba_stats <- na.omit(nba_stats)

# Drop players that were drafted before 1980
nba_stats <- nba_stats %>%
  filter(draft_year >= 1980)
```
Note: Players that were drafted before 1980 are not accounted for in order to be more indicative of the modern era.

# Marginal Summaries
```{r}
str(nba_stats)
```

```{r}
head(nba_stats)
```

```{r}
summary(nba_stats)
```

```{r}
dim(nba_stats)
```

# Questions to possibly research

## Does where you are born influence your likely hood of NBA success?
### - Could look at salary or sheer number of NBA players from certain regions and then divide by each respective population.

First I want to look at the overall salaries that are listed in this dataset to get a sort of gague of what the numbers should look like.
```{r}
#Make so there are no repeat id's and get salary statistics
nba_salary_summary <- nba_stats %>%
  group_by(player_id) %>%
  summarize(
    Total_Salary = sum(salary, na.rm = TRUE),
    Average_Salary = mean(salary, na.rm = TRUE),
    .groups = 'drop'
  )

ggplot(data = nba_salary_summary, aes(x = Total_Salary, fill = ..count..)) +
  geom_histogram(binwidth = 1000000) +
  scale_fill_gradient("Count", low = "green", high = "red") +
  scale_x_continuous(labels = label_number()) +
  labs(title = "Histogram of NBA Total Salaries",
       x = "Total Salary",
       y = "Frequency") +
  theme_minimal()

ggplot(data = nba_salary_summary, aes(x = Average_Salary, fill = ..count..)) +
  geom_histogram(binwidth = 100000) +
  scale_fill_gradient("Count", low = "green", high = "red") +
  scale_x_continuous(labels = label_number()) +
  labs(title = "Histogram of NBA Average Salaries",
       x = "Average Salary",
       y = "Frequency") +
  theme_minimal()
```
I am a little concerned that this dataset is not representative of how much players have made but since the data is pre 2019 this makes sense.

NBA salaries are increasingly inflated, especially in the last half decade, therefore it makes sense these numbers are lower than we are nowadays accustomed to. 

```{r}
library(dplyr)

# Checking for average salaries greater than $15 million
high_salary_players <- nba_salary_summary %>%
  filter(Average_Salary > 15000000)

# Viewing the results
print(high_salary_players)
count(high_salary_players)
```

Now lets look to answer "Does where you are born influence your likely hood of NBA success?"

```{r}
#Separate the birthplace column into two separate rows and look at what city/state they are from
nba_birthplace_summary <- nba_stats %>%
  distinct(player_id, .keep_all = TRUE) %>%
  separate(
    birthPlace,
    into = c("City", "State_Country"),
    sep = ", ",
    extra = "merge"
  ) %>%
  group_by(player_id, State_Country) %>%
  ungroup() %>%
  count(State_Country, sort = TRUE)

#Plot which states/countries have the most players (top 10)
N <- 10

#Create variable that shows where most players are from (state/country)
top_nba_locations <- nba_birthplace_summary %>%
  top_n(N, n) %>%
  arrange(desc(n))

#Graph Findings
ggplot(top_nba_locations, aes(
  x = reorder(State_Country, n),
  y = n,
  fill = State_Country
)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "State/Country", y = "Number of Players",
       title = "Number of NBA Players by State/Country") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Get a table that shows output
print(nba_birthplace_summary)


```

Population for each state/country
-California 39.03 M
-New York 19.68 M
-Illinois 12.58 M
-Texas 30.03 M
-Georgia 10.91 M
```{r}
#Find players per state (One NBA player every x people)
Players_CA <- 39030000 / 184 
Players_NY <- 19680000 / 117
Players_IL <- 12580000 / 92
Players_TX <- 30030000 / 68
Players_GA <- 10910000 / 65

#Expressed as a percentage
Players_CAper <- 184 / 39030000
Players_NYper <- 117 / 19680000
Players_ILper <- 92 / 12580000
Players_TXper <- 68 / 30030000
Players_GAper <- 65 / 10910000

print(Players_CA)
print(Players_NY)
print(Players_IL)
print(Players_TX)
print(Players_GA)

print(Players_CAper)
print(Players_NYper)
print(Players_ILper)
print(Players_TXper)
print(Players_GAper)

```


```{r}
#Same thing now, except group by city
nba_birthplace_city_summary <- nba_stats %>%
  distinct(player_id, .keep_all = TRUE) %>%
  separate(birthPlace, into = c("City", "State_Country"), sep = ", ", extra = "merge") %>%
  group_by(player_id, City) %>%
  ungroup() %>%
  count(City, sort = TRUE)

#Plot which cities have the most players (top 10)
top_nba_locations_cities <- nba_birthplace_city_summary %>%
  top_n(N, n) %>%
  arrange(desc(n))

#Plot findings
ggplot(top_nba_locations_cities, aes(x = reorder(City, n), y = n, fill = City)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Paired") +  # Use a pre-defined color palette
  labs(x = "City", y = "Number of Players", 
       title = "Number of NBA Players by City") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Print table
print(nba_birthplace_city_summary)
```
Populations for each city
-Los Angeles: 3.822 M
-Chicago 2.665 M
-Philadelphia 1.567 M
-Washington 671803
-Detroit 620376

```{r}
#Find players per state (One NBA player every x people)
Players_LA <- 3822000 / 54 
Players_CHI <- 2665000 / 49
Players_PHIL <- 1567000 / 34
Players_WSH <- 671803 / 31
Players_DET <- 620376 / 28

#Now express as a percentage
Players_LA_per <- 54 / 3822000
Players_CHI_per <- 49 / 2665000
Players_PHIL_per <- 34 / 1567000
Players_WSH_per <- 31 / 671803
Players_DET_per <- 28 / 620376


print(Players_LA_per)
print(Players_CHI_per)
print(Players_PHIL_per)
print(Players_WSH_per)
print(Players_DET_per)

print(Players_LA)
print(Players_CHI)
print(Players_PHIL)
print(Players_WSH)
print(Players_DET)
```



## What is the ideal height/weight (build) of an NBA Player?
- Could look at how salary relates to certain measurements a player has.
```{r}
# Create a copy of nba_stats
ns_w_cd <- nba_stats

# Add career_duration column
ns_w_cd$career_duration <- ifelse(ns_w_cd$career_G > 4*82, "Over 4 Years", "Under 4 Years")

# Filter to include only unique player IDs
unique_players <- ns_w_cd %>% 
  distinct(player_id, .keep_all = TRUE)

# Plotting the correlation between weight and career duration
weight_duration_plot <- ggplot(unique_players, aes(x = weight, y = career_G, color = career_duration)) +
  geom_point() +
  geom_smooth(aes(group = career_duration), method = "lm", se = FALSE, color = "black") +  # Add black linear regression lines
  labs(title = "Correlation between Weight and Career Duration (Unique Players)", x = "Weight", y = "Career Duration") +
  theme_minimal()

# Display the plot
print(weight_duration_plot)

# Plotting the correlation between height and career duration
height_duration_plot <- ggplot(unique_players, aes(x = height, y = career_G, color = career_duration)) +
  geom_point() +
  geom_smooth(aes(group = career_duration), method = "lm", se = FALSE, color = "black") +  # Add black linear regression lines
  labs(title = "Correlation between Height and Career Duration (Unique Players)", x = "Height", y = "Career Duration") +
  theme_minimal()

# Display the plot
print(height_duration_plot)
```

```{r}
# Boxplot of height for high-salary and low-salary players
ggplot(nba_stats, aes(x = factor(salary > median(salary)), y = height)) +
  geom_boxplot() +
  labs(title = "Height Distribution of High-Salary and Low-Salary Players",
       x = "Salary Level",
       y = "Height")

# Boxplot of weight for high-salary and low-salary players
ggplot(nba_stats, aes(x = factor(salary > median(salary)), y = weight)) +
  geom_boxplot() +
  labs(title = "Weight Distribution of High-Salary and Low-Salary Players",
       x = "Salary Level",
       y = "Weight")

```

## What are the key performance indicators for success in basketball?
```{r}
# Create a scatter plot for Position vs. Salary
ggplot(nba_stats, aes(x = position, y = salary)) +
  geom_boxplot() +
  labs(title = "Salary Distribution by Position",
       x = "Position",
       y = "Salary") +
  theme_minimal() +
  coord_flip()
```
```{r}
#Player Efficiency Rating (PER) vs. Salary
ggplot(nba_stats, aes(x = career_PER, y = salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Player Efficiency Rating (PER) vs. Salary",
       x = "Career PER",
       y = "Salary") +
  theme_minimal()
```

```{r}
#Win Shares vs. Salary
ggplot(nba_stats, aes(x = career_WS, y = salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Win Shares vs. Salary",
       x = "Career Win Shares",
       y = "Salary") +
  theme_minimal()
```
```{r}
#Draft Pick vs. Salary
ggplot(nba_stats, aes(x = draft_pick, y = salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "green") +
  labs(title = "Draft Pick vs. Salary",
       x = "Draft Pick",
       y = "Salary") +
  theme_minimal()
```
## What position in basketball seems to contribute to winning the most?
- Could look to analyze win shares and other statistics relating to positions.

```{r}
# Aggregate data by position and calculate mean career win shares
position_summary <- nba_stats %>%
  group_by(position) %>%
  summarise(mean_career_WS = mean(career_WS, na.rm = TRUE))

# Create a bar plot
ggplot(position_summary, aes(x = position, y = mean_career_WS)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average Career Win Shares by Position",
    x = "Position",
    y = "Mean Career Win Shares"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- Now look at it with just primary position
```{r}
#Make a new column that accounts for the primary position
df1$primary_position <- sapply(strsplit(df1$position, " and "), `[`, 1)

#Make sure that the number is numeric
df1$career_WS <- as.numeric(gsub("[^0-9\\.]", "", df1$career_WS))

#Group by primary_position and summarize average win shares
primary_position_summary <- df1 %>%
  group_by(primary_position) %>%
  summarise(average_win_shares = mean(career_WS, na.rm = TRUE)) %>%
  arrange(desc(average_win_shares))

#Just make sure that we only have the five basic positions
top_positions <- primary_position_summary %>%
  arrange(desc(average_win_shares)) %>%
  top_n(5, average_win_shares)

#Plotting using a bar chart
ggplot(top_positions, aes(x = reorder(primary_position, -average_win_shares), y = average_win_shares, fill = primary_position)) +
  geom_bar(stat = "identity", color = "black") +
  labs(x = "Primary Position", y = "Average Win Shares", title = "Top 5 Average Win Shares by Primary Basketball Position") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
  scale_fill_brewer(palette = "Set3")

```
